\pkg{disaggregation}: An \proglang{R} Package for Bayesian Spatial Disaggregation Modelling

### 2021-01-03 article.Rnw
Updates from review comments

%
<<kintroptions, echo=FALSE>>=
knitr::knit_hooks$set(render_sweave())
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = 'H',
  prompt=TRUE
)
options(prompt="R> ")
@
%

@@

\begin{document}

@@

Suppose we have response data, $y_i$, for $N$ polygons, which corresponds to count data for the property of interest within that polygon. 
The process that is being counted occurs in continuous space that we model as a high-resolution, square lattice for convenience.
The data, $y_i$,  are assumed to be created by the aggregation of the counts over the polygon, i.e., the count data of the polygon is given by the sum of the count data for all the pixels within that polygon. 
The rate is defined such that the number of cases in this pixel can be calculated by multiplying the rate by the aggregation raster. For example, in epidemiology, we may have as our response the number of people that contract a certain disease in a given period of time (case incidence). Our rate would be the incidence rate of cases per population, where the aggregation raster corresponds to population.

For the disaggregation model we model the rate at pixel level, with the likelihood for the observed data given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific iid effect. The user-defined link function is typically log, identity and logit for Poisson, Normal and binomial likelihoods, respectively.  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

The pixel predictions are then aggregated to the polygon level using the weighted sum (via the aggregation raster $a_{ij}$):

%
\begin{equation} \label{eq:agreggate1}
\textrm{cases}_i = \sum_{j=1}^{N_i} a_{ij}\textrm{rate}_{ij}
\end{equation}
\begin{equation} \label{eq:agreggate2}
\textrm{rate}_i = \frac{\textrm{cases}_{i}}{\sum_{j=1}^{N_i} a_{ij}}
\end{equation}
%
The different likelihoods correspond to slightly different models ($y_i$ is the response count data):

@@

For each of the model parameters and hyperparameters we specify priors. The regression parameters and intercept are given Gaussian priors, where the default priors are $\beta_0 \sim N(0, 2)$ and $\beta_i \sim N(0, 0.4)$. A penalised complexity prior is placed on the scale and range parameters of the random field \citep{fuglstad2018constructing} such that
\begin{align}
    P(\rho < \rho_{min}) &= \rho_{prob} \\
    P(\sigma > \sigma_{max}) &= \sigma_{prob}
\end{align}
where the values $\rho_{min}, \,\rho_{prob}, \,\sigma_{max}, \,\sigma_{prob}$ are set by the user. This prior shrinks the field towards a base model with zero variance and infinite range, in other words regularising towards a flatter field with smaller magnitude. The polygon-specific effects $u_1,...,u_N$ have Gaussian priors centered at 0 with standard deviation $\sigma_{u}$ (where the precision $\tau_{u} = 1/\sigma_{u}^2$). A penalised complexity prior \citep{simpson2017penalising} is placed on $\sigma_{u}$ such that 
\begin{equation}
P(\sigma_{u} > \sigma_{u, max}) = \sigma_{u, prob}
\end{equation}
which shrinks towards a base model of no polygon-specific effect. The values $\sigma_{u, max}$ and $\sigma_{u, prob}$ are set by the user.

For models that use a Gaussian likelihood, a log gamma prior is set on the log of the precision, $\log\tau \sim \log\Gamma(\text{shape} = 1, \text{rate} = 5 \times 10^{-5})$, to regularise the dispersion ($\sigma = 1/\sqrt\tau$) to take low values. This is the same as the prior set by INLA for the dispersion of the normal likelihood.

@@

@@

In this section we show how to use the \pkg{disaggregation} package using a dataset of aggregated malaria case counts across Madagascar. Malaria is an infectious disease caused by parasites of the Plasmodium group, transmitted by Anopheles mosquitoes. Malaria transmission is therefore closely related to mosquito and parasite development. Environmental factors such as temperature, rainfall and elevation have been shown to have signifcant effects on mosquito survival and development; in general mosquitoes favour warm, humid environments with moderate rainfall. Therefore, such environmental covariates would be useful in a malaria disaggregation model to inform fine-scale distributions. In this model we use the environmental covariates of mean land surface temperature (LST), elevation, and enhanced vegetation index (EVI) to inform spatial heterogeneity in malaria risk. These covariates are obtained from the Moderate Resolution Imaging Spectroradiometer (MODIS), which provides many measurements over the entire Earth's surface (\url{https://modis.gsfc.nasa.gov/data/}). 
Malaria incidence rate is often given per thousand people per year given the term annual parasite index (API).

@@

%
<<installrun, echo=FALSE, message=FALSE>>=
library(disaggregation, quietly = TRUE)
library(raster, quietly = TRUE)
@
%

@@

We then read in the data to use in the \pkg{disaggregation} package.

@@

Firstly, we use the \code{prepare_data} function to setup all the data in the format needed in the disaggregation modelling. This function performs various data manipulation tasks to create objects that are necessary for fitting the model. The required input data for the \code{prepare_data} function are a SpatialPolygonsDataFrame containing the response data and a RasterStack of covariates to be used in the model. The variable names in the SpatialPolygonsDataFrame of the response count data and the polygon ID are defined by the user.

@@

The values of the covariates (as well as the aggregation raster, if given) are extracted at each pixel within the polygons and stored as a data.frame with a row for each pixel and a column for each covariate (\code{parallelExtract} function). The extraction of each covariate is performed in parallel over the number of cores defined by the argument \code{ncores}. The values extracted from the aggregation raster are returned as an array of values, one for each pixel.

In order to know which pixels (i.e., which rows) are contained in each polygon, a matrix is constructed that contains the start and end pixel index for each polygon (\code{getStartendindex} function). Additionally, an INLA mesh is built to use for the spatial field (\code{build_mesh} function). The \code{mesh.args} argument allows the user to supply a list of INLA mesh parameters to control the mesh used for the spatial field. The mesh can take several minutes to construct, so to prepare the data without buidling the mesh the user can set the \code{makeMesh} flag to FALSE. However, it would then not be possible to fit the disaggregation model without the mesh.

@@

Now we can fit the disaggregation model using \code{disag_model}. Here we use a Poisson likelihood for the incidence count data with a log link function. Options for the likelihood are Gaussian, Poisson and binomial, and options for the link function are logit, log and identity. The spatial field and iid effect are components of the model by default, they can be turned off using the \code{field} and \code{iid} flags. In this example we use the default priors. Hyperpriors for the field are implemented as penalised complexity priors - specify $\rho_{min}$ and $\rho_{prob}$ for the range of the field, where $P(\rho < \rho_{min}) = \rho_{prob}$, and $\sigma_{max}$ and $\sigma_{prob}$ for the variation of the field, where $P(\sigma > \sigma_{max}) = \sigma_{prob}$. Similarly, penalised complexity priors are implemented for the iid effect.
The \code{iterations} argument specifies the maximum number of iterations the model can run for to find an optimal point. In order to print more verbose output the user can set the \code{silent} argument to FALSE.

%
<<disag_model, message = FALSE>>=
fitted_model <- disag_model(data = dis_data,
                            iterations = 1000,
                            family = 'poisson',
                            link = 'log')
@
%

We can get a summary and plot of the model output. The \code{summary} function gives the estimate and standard error of the fixed effect parameters in the model, as well as the negative log likelihood and in-sample performance metrics (root mean squared error, mean absolute error, pearson correlation coefficient and spearman rank correlation coefficient). The \code{plot} function produces two plots: one of the fixed effects parameters and one of the observed data against in-sample predictions, as shown in Figure~\ref{fig:plotmodel}.

@@

The \code{plot} function can be used on the \code{disag_prediction} objects, as shown in Figure \ref{fig:plotprediction}.

@@

@@

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS (no-u-turn sampler) algorithm \citep{hoffman2014no} using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{disag_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 71 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations, doubling the number of iterations each time until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. In contrast, fitting the model using the Laplace approximation via \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}.

>>>>

%
<<kintroptions, echo=FALSE>>=
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = 'H',
  prompt=TRUE
)
options(prompt="R> ")
@
%

@@

\begin{document}
%\SweaveOpts{concordance=TRUE}

@@

Suppose we have response data, $y_i$, for $N$ polygons, which corresponds to count data for the property of interest within that polygon. 
The process that is being counted occurs in continuous space that we model as a high-resolution, square lattice for convenience.
The data, $y_i$, are assumed to be created by the aggregation of the counts over the polygon, i.e., the count data of the polygon is given by the sum of the count data for all the pixels within that polygon. As well as the model covariates, an aggregation raster can be provided transform from the pixel level predictions (rate) to count data, from which it is trivial to calculate the corresponding polygon value by summing up the values for the individual pixels within the polygon. For example, in epidemiology, we may have as our response the number of people that contract a certain disease in a given period of time (case incidence). Our rate would be the incidence rate of cases per population, where the aggregation raster corresponds to population. Predictions of incident rate per population are made at pixel level, the population (aggregation) raster is used to convert these values to number of cases, which can then be summed to get predicted polygon level cases.

For the disaggregation model we model the rate at pixel level, with the likelihood for the observed data given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific iid effect. The user-defined link function is typically log, identity and logit for Poisson, Normal and binomial likelihoods, respectively.  The Gaussian random field has a Matern covariance function, defined by:

%
\begin{equation}
Cov(d) = \frac{\sigma^2}{\Gamma(\nu)2^{\nu - 1}}(\kappa d)^{\nu}K_{\nu}(\kappa d)
\end{equation}
%

with two hyperparameters: $\rho = \frac{\sqrt{8\nu}}{\kappa}$, the range of the field (beyond which correlation is $<$ 0.1), and $\sigma$, the marginal standard deviation, relating to the magnitude of the variation in the field. The parameter $\nu$ is the smoothness and is fixed at 2. The Matern field is at least $m$ times mean square differentiable if and only if $\nu > m$~\citep{stein1999interpolation}, therefore $\nu = 2$ is chosen as it is the smallest integer (least smooth) for the field to be differentiable. The Gamma function is given by $\Gamma(\nu) = (\nu - 1)!$ and $K_{\nu}$ is the modified Bessel function of the second kind.

As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

The predictions are made at pixel level, however the response data exists as count data at the polygon level. Therefore, to calculate the contribution  to the likelihood from the data we must aggregate the pixel predictions, (via the aggregation raster $a_{ij}$, to get the predicted polygon count.

%
\begin{equation} \label{eq:agreggate1}
\textrm{cases}_i = \sum_{j=1}^{N_i} a_{ij}\textrm{rate}_{ij}
\end{equation}
\begin{equation} \label{eq:agreggate2}
\textrm{rate}_i = \frac{\textrm{cases}_{i}}{\sum_{j=1}^{N_i} a_{ij}}
\end{equation}
%
where $N_i$ is the total number of pixels in polygon $i$.
The different likelihoods correspond to slightly different models ($y_i$ is the response count data):

@@

For each of the model parameters and hyperparameters we specify priors. The regression parameters and intercept are given Gaussian priors, where the default priors are $\beta_0 \sim N(0, 2)$ and $\beta_i \sim N(0, 0.4)$. It is expected that some of the spatial variation can be described by the covariates, and the Gaussian field can help describe the remaining spatial variation that is missing from the covariate information. Therefore, the priors on the covariates can be set to allow the covariates to explain some, but not all, of the variation in the response data. For the Gaussian random field, penalised complexity priors are used, which are constructed to  penalise against deviating from the simpler base model, which in this case is a flat field, i.e. zero variance and infinite range~\citep{fuglstad2018constructing}. A penalised complexity prior is placed on the scale and range parameters of the random field such that
\begin{align}
    P(\rho < \rho_{min}) &= \rho_{prob} \\
    P(\sigma > \sigma_{max}) &= \sigma_{prob}
\end{align}
where the values $\rho_{min}, \,\rho_{prob}, \,\sigma_{max}, \,\sigma_{prob}$ are set by the user. The default values for these parameters within the package are driven by the nature of the data provided in the model. The default prior for $\rho_{min}$ is set at a third of the spatial area covered by the polygons, and $\rho_{prob}$ is set at 0.1. The default prior for $\sigma_{max}$ is set to the standard deviation of the normalised response data, and $\sigma_{prob}$ is set at 0.1.

The joint penalised complexity prior, $\pi$, corresponding to a base model with infinite range and zero variance~\citep{fuglstad2018constructing} is given by:

%
\begin{equation} 
log\left(\pi(\sigma, \rho)\right) = log\left(\tilde{\lambda_1}\right) +  log\left(\tilde{\lambda_2}\right) - 2log(\rho) - \frac{\tilde{\lambda_1}}{\rho} - \tilde{\lambda_2} \sigma
\end{equation}

where

\begin{equation} 
\tilde{\lambda_1} = -\rho_{min}log(\rho_{prob}) \qquad \text{and} \qquad \tilde{\lambda_2} = -\frac{log(\sigma_{prob})}{\sigma_{max}} 
\end{equation}
%

This prior shrinks the field towards a base model with zero variance and infinite range, in other words regularising towards a flatter field with smaller magnitude.

The polygon-specific effects $u_1,...,u_N$ have Gaussian priors centered at 0 with standard deviation $\sigma_{u}$ (where the precision $\tau_{u} = 1/\sigma_{u}^2$). A penalised complexity prior is placed on $\sigma_{u}$~\citep{simpson2017penalising} such that 
\begin{equation}
P(\sigma_{u} > \sigma_{u, max}) = \sigma_{u, prob}
\end{equation}
where values $\sigma_{u, max}$ and $\sigma_{u, prob}$ are set by the user.
The penalised complexity prior, $\pi_u$, corresponding to a base model with no polygon-specific effect~\citep{simpson2017penalising} is given by:

%
\begin{equation} 
log\left(\pi(\lambda)\right) = log\left(\frac{\lambda}{2}\right) - \frac{3}{2}log(\tau_u) - \frac{\lambda}{\sqrt{\tau_u}} 
\end{equation}

where

\begin{equation} 
\lambda = -\frac{log(\sigma_{u, prob})}{\sigma_{u, max}} \qquad \text{and} \qquad \tau_u = \frac{1}{\sigma_u^2}
\end{equation}
%

This prior shrinks towards a base model of no polygon-specific effect.

For models that use a Gaussian likelihood, a log gamma prior is set on the log of the precision, $\log\tau_u \sim \log\Gamma(\text{shape} = 1, \text{rate} = 5 \times 10^{-5})$, to regularise the dispersion, $\sigma_u$, to take low values. This is chosen to be consistent with the prior set by INLA for the dispersion of the normal likelihood.

@@

\pkg{TMB} can fit both Frequentist and Bayesian models. In this case we have chosen the Bayesian approach as it is most widely used in the spatial modelling community.

@@

In this section we show how to use the \pkg{disaggregation} package using a dataset of aggregated malaria case counts across Madagascar in 2016. Malaria is an infectious disease caused by parasites of the Plasmodium group, transmitted by Anopheles mosquitoes. Malaria transmission is therefore closely related to mosquito and parasite development. Environmental factors such as temperature, rainfall and elevation have been shown to have signifcant effects on mosquito survival and development; in general mosquitoes favour warm, humid environments with moderate rainfall. Therefore, such environmental covariates would be useful in a malaria disaggregation model to inform fine-scale distributions. In this model we use the environmental covariates of mean land surface temperature (LST), elevation, and enhanced vegetation index (EVI) to inform spatial heterogeneity in malaria risk. These covariates are obtained from the Moderate Resolution Imaging Spectroradiometer (MODIS), which provides many measurements over the entire Earth's surface (\url{https://modis.gsfc.nasa.gov/data/}).

Malaria incidence rate is often given per thousand people per year given the term annual parasite index (API).
In this case our pixel predictions correspond to malaria incidence rate, so we use population to aggregate pixel incidence rate by summing the number of cases (rate weighted by population). Raster surfaces of population for the years 2010 and 2015, were created using data from WorldPop~\citep{tatem2017worldpop} and from GPWv4~\citep{GpwNasa} where WorldPop did not have values. The population raster for 2016 was created by linear interpolation between 2010 and 2015 and extending out to 2016. This interpolation method results in a small amount of uncertainty in population raster, however this is expected to be negligible compared to the model uncertainty.

Covariate rasters or population rasters may be accompanied by significant uncertainty. There is no established method within the \pkg{disaggregation} package to deal with this. In order to propagate this uncertainty to the predictions, users may run the model many times, varying these rasters within the known uncertainty. Both the range in predictions for different models and the model uncertainty within a model allows a quantification of the overall uncertainty of predictions.

The covariate rasters and aggregation raster provided must be of the same spatial scale and must be spatially aligned. For this example the rasters used are at a resolution of approximately $5\times5~km$ across Madagascar. For spatially misaligned rasters, there are packages in \prolang{R} to align rasters such as \code{align_rasters} in the \pkg{gdalUtils} package~\cite{gdalutils}.

@@

%
<<installrun, echo=FALSE, message=FALSE, warnings=FALSE>>=
library(disaggregation, quietly = TRUE)
library(raster, quietly = TRUE)
@
%

@@

We then read in the data to use in the \pkg{disaggregation} package. The \code{shapefile} function and \code{raster} function are contained within the \pkg{raster} package, they are functions to read spatial data and raster data respectively. These functions return objects of type SpatialPolygonsDataFrame and RasterLayer respectively. A SpatialPolygonsDataFrame is an \prolang{R} class that holds spatial data in the form of polygons with attributes, in this case the attributes are number of malaria cases within the polygon and polygon ID. A RasterLayer is an \prolang{R} class that holds a single raster. The function \code{getCovariateRasters} is a helper function contained within the \pkg{disaggregation} package. The user provides a directory name and shapefile, it finds all the raster files within that directory and creates a RasterStack with the same extend as the shapefile provided. In this example we are building a RasterStack of the covariate rasters with the same extent as the population raster. A RasterStack is a collection of RasterLayer objects with the same spatial extent and resolution.

@@

Firstly, we use the \code{prepare_data} function to setup all the data in the format needed in the disaggregation modelling. This function performs various data manipulation tasks to create objects that are necessary for fitting the model. The required input data for the \code{prepare_data} function are:
\begin{itemize}
\item \code{polygon_shapefile}: SpatialPolygonsDataFrame containing the response data. It must contain IDs and response data.
\item \code{covariate_raster}: RasterStack of covariates to be used in the model.
\item \code{aggregation_raster}: RasterLayer used as the weights to aggregate the pixel values within a polygon.
\item \code{mesh.args}: List of parameters to control the mesh used for the Gaussian Field component.
\item \code{id_var}: the variable name of the ID variable in the \code{polygon_shapefile}.
\item \code{response_var}: the variable name of the response variable in the \code{polygon_shapefile}.
\item \code{na.action}: Boolean. Whether to deal with NAs or not.
\item \code{ncores}: Number of cores to perform the parallel extraction over.
\end{itemize}

@@

The values of the covariates (as well as the aggregation raster, if given) are extracted at each pixel within the polygons and stored as a data.frame with a row for each pixel and a column for each covariate (\code{parallelExtract} function). The extraction of each covariate is performed in parallel over the number of cores defined by the argument \code{ncores}. The values extracted from the aggregation raster are returned as an array of values, one for each pixel. In order to know which pixels (i.e., which rows) are contained in each polygon, a matrix is constructed that contains the start and end pixel index for each polygon (\code{getStartendindex} function).

To fit a model with a Gaussian field we approximate a Gaussian field using Gaussian Random Markov Fields (GRMF), and solve using the stochastic partial differential equation (SPDE) approach~\citep{Lindgren2011}. This reduces the complexity of the problem to a sparse precision matrix, allowing explicit computation of the GMRF representation. This approach requires building a mesh, i.e. splitting the space into finite elements, and calculating field values at the mesh nodes. If a mesh is too fine, the field values take too long to compute, however if the mesh if too coarse, the resulting field is a poor approximation and leads to mesh artifacts in the predictions. The 2D mesh for the spatial field is built using the \code{build_mesh} function, which makes use of the \pkg{INLA} function \code{inla.mesh.2d}. The user can control the parameters of the mesh, including the granularity, using the argument \code{mesh.args} in the \code{build_mesh} function. The parameters \code{max.edge}, \code{cut} and \code{offset} can be set in the \code{mesh.args} list and are defined within the \code{inla.mesh.2d} function. By providing a two values to the \code{max.edge} parameter, the mesh contains an inner region of finer mesh, and an outer coarse region. This approach allows the region of interest to have a fine a mesh as necessary without also requiring time consuming computations in the sea regions. The mesh can take several minutes to construct, so to prepare the data without building the mesh the user can set the \code{makeMesh} flag to FALSE. However, it would then not be possible to fit the disaggregation model without the mesh.

@@

Now we can fit the disaggregation model using \code{disag_model}. The required inputs for the \code{disag_model} function are:
\begin{itemize}
\item \code{data}: Object of class \code{disag_data} returned by the \code{prepare_data} function.
\item \code{priors}: List of priors to use for the model. For any not set, the default priors will be used.
\item \code{iterations}: Maximum number of iterations the model can run for to find an optimal point.
\item \code{family}: Likelihood function. Options are gaussian, poisson and binomial.
\item \code{link}: Link function used in the model. Options are logit, log and identity. This would typically be log, identity and logit for Poisson, Normal and binomial likelihoods, respectively.
\end{itemize}

Here we use a Poisson likelihood for the incidence count data with a log link function. The spatial field and iid effect are components of the model by default, they can be turned off using the \code{field} and \code{iid} flags. In this example we use the default priors. Hyperpriors for the field are implemented as penalised complexity priors - specify $\rho_{min}$ (\code{prior_rho_min}) and $\rho_{prob}$ (\code{prior_rho_prob}) for the range of the field, where $P(\rho < \rho_{min}) = \rho_{prob}$, and $\sigma_{max}$ (\code{prior_sigma_max}) and $\sigma_{prob}$ (\code{prior_sigma_prob}) for the variation of the field, where $P(\sigma > \sigma_{max}) = \sigma_{prob}$. Similarly, penalised complexity priors are implemented for the iid effect (\code{prior_iideffect_sd_max} and \code{prior_iideffect_sd_prob}). Any of the priors and hyperpriors can be set in the list of priors given to the \code{priors} argument of the \code{disag_model} function. In order to print more verbose output the user can set the \code{silent} argument to FALSE.

%
<<disag_model, message = FALSE, warnings=FALSE>>=
fitted_model <- disag_model(data = dis_data,
                            iterations = 1000,
                            family = 'poisson',
                            link = 'log')
@
%

We can get a summary and plot of the model output. The \code{summary} function gives the estimate and standard error of the fixed effect parameters in the model, as well as the negative log likelihood and in-sample performance metrics (root mean squared error, mean absolute error, pearson correlation coefficient and spearman rank correlation coefficient). The standard error is calculated using the delta method~\citep{doob1935}, via the \code{sdreport} method in the \pkg{TMB} package. The \code{plot} function produces two plots: one of the fixed effects parameters and one of the observed data against in-sample predictions, as shown in Figure~\ref{fig:plotmodel}.

@@

The \code{plot} function can be used on the \code{disag_prediction} objects, as shown in Figure \ref{fig:plotprediction}. From Figure~\ref{fig:plotdata}, it can be seen that the polygon response data does not include the islands, whereas the covariate rasters do. As can be seen in Figure~\ref{fig:plotprediction}, out-of-sample malaria predictions of these islands have been made, however these predictions do not contribute to the in-sample performance in Figure~\ref{fig:plotmodel}.

@@

The same technique can be used for spatio-temporal disaggregation modelling. In order to achieve this, dynamic covariates are required, for example, annual covariate rasters, as well as a spatio-temporal field, which could be achieved by adding a time varying autoregressive component to the spatial field. A difficulty with this approach is that it results in significant increase in computational complexity of the optimisation problem, most notably from the spatio-temporal field, leading to a significant increase in the time taken to perform the optimisation. The package can be developed to include spatio-temporal disaggregation models, however that is beyond the scope of this paper.

@@

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS (no-u-turn sampler) algorithm \citep{hoffman2014no} using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{disag_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 71 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations, doubling the number of iterations each time until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. The property $\hat{R}$ is known as the Gelman-Rubin diagnostic~\citep{Rhat}, and is commonly used to evaluate MCMC convergence across multiple chains. In contrast, fitting the model using the Laplace approximation via \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}. It can be seen that the model parameters are very similar between the two methods. It is worth noting that the field parameters given in the Table are $log(\rho)$ and $log(\sigma)$. If we transform these parameters, the values given in Table~\ref{tab:mcmcresults} correspond to a difference in the range of the field from 1.9 (TMB) to 2.3 (MCMC), and a difference in the standard deviation of the field from 1.0 (TMB) and 1.06 (MCMC). Considering this along with the large accompanying uncertainties, these field parameters are very similar.
